---
title: "Project proposal"
author: "Approximately Normal: Jenny Huang, Bella Larsen, Jessie Bierschenk, Aidan Gildea"
date: "6 October 2020"
output: pdf_document
---

```{r setup, echo = F}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE, 
                      fig.align = "center")
```

```{r load-packages, message=FALSE}
library(tidyverse)
library(broom)
library(patchwork)
```

```{r data}
spotify <- read_csv("data/spotify_songs.csv")
```

### Section 1. Introduction

### Section 2. Data description

```{r obs}
observations <- as.tibble(count(spotify)) %>%
  mutate(variables = ncol(spotify))
observations %>%
  kable()
```
In our data set `spotify`, there are 32,833 observations, each of which is a song
on Spotify. There are 23 variables in the data set, ranging from the song's release
date to its danceability. 

```{r response}
spotify[,c("playlist_genre", "playlist_subgenre")] 
```
The primary response variable of our regression analysis will be 
`playlist_genre`. Using the other characteristic variables of the songs
('key', 'loudness', 'acousticness, etc.), we will predict the genre of the 
playlist each song would be placed in. Depending on the criteria of the project,
we may also extend our analysis to predict the subgenre of the playlist a
song is in based off its characteristics.

### Section 3. Glimpse of data 

```{r glimpse}
glimpse(spotify)
```


