---
title: "Your project title"
author: "Approximately Normal: Jenny Huang, Bella Larsen, Jessie Bierschenk, Aidan Gildea"
date: "28 October 2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
```

```{r load-packages, message=FALSE}
library(tidyverse)
library(broom)
library(patchwork)
library(RColorBrewer)
library(ggthemes)
library(viridis)
```

```{r data}
elections <- read_csv("data/data_full.csv")
```

```{r data cleaning}
elections_clean <- elections %>% 
  mutate(metro = if_else(METRO == 2, "Metro", "Not Metro/Unknown")) %>% 
  mutate(sex = case_when(SEX == 1 ~ "Male",
                         SEX == 2 ~ "Female")) %>% 
  mutate(marst = case_when(MARST == 1 ~ "Married",
                           MARST == 2 ~ "Married",
                           MARST == 4 ~ "Divorced/Separated",
                           MARST == 3 ~ "Divorced/Separated",
                           TRUE ~ "Not Married/Other")) %>% 
  mutate(veteran = if_else(VETSTAT == 2, "Yes", "No/Uknnown")) %>% 
  mutate(citizen = case_when(CITIZEN == 5 | CITIZEN == 9 ~ "No/Unknown", 
                             CITIZEN == 1 | CITIZEN == 2 | CITIZEN == 3 ~ 
                               "Native Born",
                             CITIZEN == 4 ~ "Naturalized")) %>% 
  #We should probably exclude people who are not citizens from our analysis 
  #because they are not able to vote
  mutate(hispanic_status = if_else(HISPAN == 0 | HISPAN == 901 | HISPAN == 902, 
                             "Not Hispanic/Unknown", "Hispanic/Latinx")) %>% 
  mutate(employed = if_else(LABFORCE == 2, "Yes", "No/Unknown")) %>% 
  mutate(highest_education = case_when(EDUC99 == 0 | EDUC99 == 1 ~ 
                                         "None/Unknown",
                               EDUC99 == 4 | EDUC99 == 5 | EDUC99 == 6 |
                                 EDUC99 == 7 | EDUC99 == 8 | EDUC99 == 9
                               ~ "Some High School",
                               EDUC99 == 10 ~ "High School Degree/GED",
                               EDUC99 == 11 ~ "Some College",
                               EDUC99 == 12 | EDUC99 == 13 | EDUC99 == 14
                               ~ "Associate Degree",
                               EDUC99 == 15 ~ "Bachelors Degree",
                               EDUC99 == 16 ~ "Masters Degree",
                               EDUC99 == 17 ~ "Professional Degree",
                               EDUC99 == 18 ~ "Doctorate Degree")) %>% 
  mutate(current_student = case_when(SCHLCOLL == 5 | SCHLCOLL == 0 ~ 
                                       "Not currently enrolled",
                                     SCHLCOLL == 1  ~ "High School Full Time",
                                     SCHLCOLL == 2 ~ "High School Part Time",
                                     SCHLCOLL == 3 ~ "College Full Time",
                                     SCHLCOLL == 4 ~ "College Part Time")) %>% 
#of people 16-24
  mutate(race = case_when(RACE == 820 | RACE == 830 | RACE == 830 | RACE == 819 
                          | RACE == 804 | RACE == 805 | RACE == 806 | 
                            RACE == 807 | RACE == 808 | RACE == 810 | 
                            RACE == 811 | RACE == 812 | RACE == 813 | 
                            RACE == 814 | RACE == 815 | RACE == 816 | 
                            RACE == 817 | RACE == 818 | RACE == 803 | 
                            RACE == 801 | RACE == 802 | RACE == 830 ~ 
                            "2 or more races",
                          RACE == 100 ~ "White",
                          RACE == 200 ~ "Black",
                          RACE == 651 | RACE == 809 | RACE == 652 | RACE == 650
                          ~ "Asian or Pacific Islander",
                          RACE ==  300 ~ "Native American",
                          RACE == 999 | RACE == 700 ~ "Other/Unknown")) %>% 
  #Need to go over these race categorizations as a group!
  mutate(why_not_vote = case_when(VOWHYNOT == 10 ~ "Inconvenience",
                                  VOWHYNOT == 4 ~ "Interest",
                                  VOWHYNOT == 7 ~ "Political",
                                  VOWHYNOT == 9 | VOWHYNOT ==  6 | 
                                    VOWHYNOT == 5 | VOWHYNOT == 2 ~ 
                                    "Logistical",
                                  VOWHYNOT == 1 ~ "Physically Unable",
                                  VOWHYNOT == 8 ~ "Registration Issues",
                                  VOWHYNOT == 3 ~ "Forgot")) %>% 
#Reason why eligible voter did not vote
  mutate(why_not_reg = case_when(VOYNOTREG == 8 | VOYNOTREG == 3 ~ 
                                   "Not Eligible",
                                 VOYNOTREG == 7 | VOYNOTREG == 6 ~ 
                                   "Not Interested/Vote Won't Matter",
                                 VOYNOTREG == 5 ~ "Language Barrier",
                                 VOYNOTREG == 4 ~ "Physically Unable",
                                 VOYNOTREG == 2 | VOYNOTREG == 1 ~ 
                                   "Lacked Info/Missed Deadline")) %>% 
  mutate(voting_method = case_when(VOTEHOW == 1 ~ "In Person",
                                   VOTEHOW == 2 ~ "Mail-In")) %>% 
  mutate(voting_time = case_when(VOTEWHEN == 2 ~ "Early",
                                 VOTEWHEN == 1 ~"Voting Day")) %>% 
  mutate(voted = case_when(VOTED == 1 ~ 0,
                           VOTED == 2 ~ 1)) %>% 
  mutate(registered = case_when(VOREG == 1 ~ 0,
                           VOREG == 2 ~ 1)) %>% 
  mutate(how_registered = case_when(VOREGHOW == 8 ~ "Online",
                                    VOREGHOW == 7 ~ "Polling Place",
                                    VOREGHOW == 6 ~ "Registration Drive",
                                    VOREGHOW == 5 | VOREGHOW == 2 | 
                                      VOREGHOW == 1 ~ 
                                      "Govt Office/Public Agency",
                                    VOREGHOW == 4 ~ "School/College/Hospital",
                                    VOREGHOW == 3 ~ "By Mail")) %>% 
  select(metro, sex, marst, veteran, citizen, hispanic_status, employed, 
         highest_education, current_student, race, why_not_vote, why_not_reg,
         voting_method, voting_time, voted, registered, how_registered, YEAR,
         STATEFIP, AGE)

elections_clean <- elections_clean %>% 
  mutate(metro = factor(metro),
         sex = factor (sex),
         marst = factor(marst),
         veteran = factor(veteran),
         citizen = factor(citizen),
         hispanic_status = factor( hispanic_status),
         employed = factor(employed),
         highest_education = factor(highest_education),
         current_student = factor(current_student),
         race = factor(race),
         why_not_vote = factor(why_not_vote),
         why_not_reg = factor(why_not_reg),
         voting_method = factor(voting_method),
         voting_time = factor(voting_time),
         voted = factor(voted),
         registered = factor(registered),
         how_registered = factor(how_registered))

```

### Introduction

We will begin our EDA by visualizing the relationship between the response variable
`voted` and several of the other variables of particular interest.

We will begin by simply looking at the distribution of those who voted throughout 
the last 8 years of elections.

```{r EDA}
response_plot <- ggplot(data = elections_clean, aes(x = voted, fill = "gainsboro")) +
  geom_bar() +
  theme_bw() +
  # see theme code inspiration at reference 1 
  theme(plot.title= element_text(face="bold", color = "Navy Blue"),
        plot.subtitle = element_text(face = 4, color = "grey54"),
        panel.grid.major = element_line(color = "Navy Blue", size=.1),
        legend.position = "none") +
  # see scale fill code inspiration at reference 2
  scale_fill_manual(values = c("gainsboro")) +
  labs(title = "Visualizing the Distribution of Voting Status", y = "Count",
       x = "Voted", subtitle = "More people reportedly voted than did not vote",
       caption = "0 = Did not Vote, 1 = Voted")
response_plot
```

From the barplot above, it is clear the more individuals in the data set voted (voted = 1) than did not (voted = 0).

As college students ourselves, we want to analyze whether or not being a student influences the frequency of voting. We will explore this preliminarily by visualizing the distribution of if school aged individuals (18-24) voted or not -- categorized by their current student level. This is seen in the bar plot below.

```{r}
EDA1 <- elections_clean %>% 
  ggplot(aes(x = voted, fill = current_student)) +
  geom_bar() +
  theme_bw() +
  theme(plot.title= element_text(face="bold", color = "Navy Blue"),
        plot.subtitle = element_text(face = 4, color = "grey54"),
        panel.grid.major = element_line(color = "Navy Blue", size=.1))+
  #see scale fill brewer code inspiration from reference 1
  scale_fill_brewer(palette = "Blues") +
  labs(title = "Voting Distribution of Population of 16-24 Year Olds", 
       y = "Count", x = "Voted", 
       subtitle = "Examining relationship between student status and voting",
       caption = "0 = Did not Vote, 1 = Voted",
       fill = "Current Student Status")
EDA1
```

From the bar plot, it is evident that a majority of these individuals were not currently enrolled. This may be a result of a general national trend, but we want to investigate if it is the result of a larger proportion of older individuals within in the range of ages between 16-24. We will investigate this by analyzing those who are not currently enrolled in school within this age range.

```{r}
elections_clean %>%
  filter(current_student == "Not currently enrolled") %>%
  filter(AGE >=16 & AGE <=24) %>%
  group_by(AGE) %>%
  count()  %>% 
  ungroup() %>%
  mutate(prop = n/sum(n)) %>%
  kable(digits = 3)
```
From the kable above, it is apparent that more than 40% of those not currently enrolled in school are 23-24 years old. This could be a potential reason for why this age range includes so many who are not currently enrolled as a student. 

To more meaningfully analyze the relationship between being a student and if they vote or not, we adjusted our visualization to only include those currently enrolled in some level of education. This is seen in the visualization below. 

```{r}
EDA2 <- elections_clean %>% 
  filter(current_student != "Not currently enrolled") %>% 
  ggplot(aes(x = voted, fill = current_student)) +
  geom_bar() +
  theme_bw() +
    theme(plot.title= element_text(face="bold", color = "Navy Blue"),
        plot.subtitle = element_text(face = 4, color = "grey54"),
        panel.grid.major = element_line(color = "Navy Blue", size=.1))+
  scale_fill_brewer(palette="Blues") +
  labs(title = "Voting Distribution of Population of 16-24 Year Olds 
Enrolled in School", 
       y = "Count", x = "Voted", 
       subtitle = "Examining relationship between student status and voting",
       caption = "0 = Did not Vote, 1 = Voted",
       fill = "Current Student Status")
EDA2
```

```{r EDA3}
elections_clean %>% 
  ggplot(aes(x = voted, fill = citizen)) +
  geom_bar() +
  scale_fill_brewer(palette="Paired") +
  labs(title = "Voting distribution based on citizenship status", 
       y = "Count", x = "Voted", 
       subtitle = "Examining relationship between student status and voting",
       caption = "0 = Did not Vote, 1 = Voted",
       fill = "Citizenship Status")

```

```{r EDA4}
elections_clean %>% 
  ggplot(aes(x = AGE, y = voted)) +
  geom_boxplot() +
  coord_flip()+
  labs(title = "Relationship between age and voting", 
       y = "Voting Status", x = "Age", 
       caption = "0 = Did not Vote, 1 = Voted",
       fill = "Citizenship Status")

```

```{r EDA5}
elections_clean %>% 
  ggplot(aes(x = voted, fill = veteran)) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette="Paired") +
  labs(title = "Voting distribution based on veteran status", 
       y = "Count", x = "Voted", 
       subtitle = "Examining relationship between veteran status and voting",
       caption = "0 = Did not Vote, 1 = Voted",
       fill = "Veteran Status")

```
We are also interested in looking at how voter turnout has changed over the
years.

```{r year-voting-frequency}
elections_year <- elections_clean %>% 
  group_by(YEAR) %>% 
  count(voted) %>% 
  mutate(prop_voted = n/sum(n))
```

We notice that the proportion of people who voted fluctuates depending on
whether the year falls on a presidential election. In the trend of the 
proportion of voting over time, we see a clear divide between the years when
there is a presidential elections versus when there is not. In the future, we 
may decide to add the variable "Election Year" as an interaction term with year
as a divide between years that fall on an election.

```{r EDA6}
elections_year %>% 
  filter(voted == 1) %>% 
  ggplot(aes(x = YEAR, y = prop_voted)) +
  geom_point() +
  labs(title = "More Voters During Presidential Election Years", 
       y = "Proportion of People Who Voted", x = "Year")
```

```{r new-variable}
elections_clean <- elections_clean %>% 
  mutate(Presidential_Election_Year = if_else(YEAR %in% c(2004, 2008, 
                                                          2012, 2016), 1, 0),
         Presidential_Election_Year = factor(Presidential_Election_Year))
```

### Model Selection 

```{r model-selection}

elections_model <- elections_clean%>%
  filter(YEAR==2016) 

full_model <- glm(voted ~ metro + sex + marst + veteran+ citizen + 
                   hispanic_status + employed + highest_education + 
                   current_student + race + AGE,
                     data = elections_model,
                      family = "binomial")
Backward_Model <- step(full_model, direction = "backward")
tidy(Backward_Model)

int_only_model <- glm(voted ~1,
                     data = elections_model,
                      family = "binomial")
Backward_Model2 <- step(full_model, scope=formula(int_only_model), direction = "backward")
tidy(Backward_Model2)


```


```{r model-selection-test}
elections_model = elections_clean%>%
  filter(YEAR==2016)
full_model = glm(voted ~ metro + marst + race + AGE,
                     data = elections_model,
                      family = "binomial")
Backward_Model <- step(full_model, direction = "backward")
tidy(Backward_Model)

int_only_model = glm(voted ~1,
                     data = elections_model,
                      family = "binomial")
Backward_Model2 <- step(full_model, scope=formula(int_only_model), direction = "backward")
tidy(Backward_Model2)
```
### Checking Model Conditions 


[1]: http://www.sthda.com/english/articles/32-r-graphics-essentials/125-ggplot-cheat-sheet-for-great-customization/#use-themes-in-ggplot2-package
[2]: https://www.datanovia.com/en/blog/the-a-z-of-rcolorbrewer-palette/

keep in mind: citizenship and registration exclusion for the model